Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach

### 2014-05-13 manuscripts/manuscript.Rmd

Decision making under uncertainty is a ubiquitous challenge of
natural resource management and conservation.  Ecological dynamics
are frequently complex and difficult to measure, making uncertainty
in our understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Concerns
about the potential sudden collapse of fisheries and other ecosystems
[@Worm2006, @Scheffer2009] has increasingly emphasized the difficulties of
formulating even qualitatively correct models of the underlying processes.

@@

We use a Bayesian nonparameteric regression approach based on Gaussian process priors. 
The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
GP models have subsequently been used to test for the presence of Allee
effects [@Sugeno2013], estimate the maximum reproductive
rate [@Sugeno2013], determine temporal variation in food
availability [@Sigourney2012], and provide a basis for identifying
model-misspecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006. The nonparametric moniker for
the GP approach is unfortunate as (a) it still involves the estimation
of parameters, and (b), statisticians also use non-parametric to mean
models that make only weak distributional assumptions  [e.g. @Lehmann1975]. 
However, we retain the terminology for consistency with previous
studies.

@@

We inferred posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (@Gelman2003 implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). We show One-step-ahead predictions of these model fits in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results were robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

```{r Figure_5, dependson=c("sim"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model."}
ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=.1) +
  scale_colour_manual(values=colorkey, guide = guide_legend(override.aes = list(alpha = 1))) + 
  facet_wrap(~method) + guides(legend.position="none")

@@

@@

Lastly, it should be noted that outside the data, the GP reverts to
the prior, and consequently the choice of the prior can also play a
significant role in determining the optimal policy inferred by the
SDP.  In the examples shown here we have selected a prior that is both
relatively uninformative (due to the broad priors placed on its parameters
$\ell$ and $\sigma$ and simple (mean zero, radial basis function kernel).
In practice, both the choice of mean and the choice of the covariance
function may be chosen to confer particular biological properties,
as well as more biologically informed priors for $\ell$ and $\sigma$.
In principle, this may allow a manager to improve the performance of the
GPDP approach by adding only enough additional detail as is justified.
For instance, it would be possible to use a linear or a Ricker-shaped
mean in the prior without making the much stronger assumption that the
Ricker is the structurally correct model [@Sugeno2013]. Future research
should focus on identifying criteria to ensure the prior and the value function
are chosen appropriately for the problem at hand.

>>>>

Decision making under uncertainty is a ubiquitous challenge of
natural resource management and conservation.  Ecological dynamics
are frequently complex and difficult to measure, making uncertainty
in our understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Concerns
about the potential sudden collapse of fisheries and other ecosystems
[@Worm2006; @Scheffer2009] has increasingly emphasized the difficulties of
formulating even qualitatively correct models of the underlying processes.

@@

We use a Bayesian nonparameteric regression approach based on Gaussian process priors. 
The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
GP models have subsequently been used to test for the presence of Allee
effects [@Sugeno2013a], estimate the maximum reproductive
rate [@Sugeno2013b], determine temporal variation in food
availability [@Sigourney2012], and provide a basis for identifying
model-misspecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006. The nonparametric moniker for
the GP approach is unfortunate as (a) it still involves the estimation
of parameters, and (b), statisticians also use non-parametric to mean
models that make only weak distributional assumptions  [e.g. @Lehmann1975]. 
However, we retain the terminology for consistency with previous
studies.

@@

We inferred posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (@Gelman2003 implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See appendix Tables S1-S3; R code
provided). We show one-step-ahead predictions of these model fits in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results were robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

```{r Figure5, dependson=c("sim"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model." fig.width=6, fig.height=6}
ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=.1) +
  scale_colour_manual(values=colorkey, guide = guide_legend(override.aes = list(alpha = 1))) + 
  facet_wrap(~method) + guides(legend.position="none")

@@

Large noise relative to the distance between the stable and unstable
point, (or more precisely, between the optimal escapement and the unstable
point) increases the chance of a stochastic transition.  As a result,
the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be modeled
near-optimally. Models which underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficently rare.

@@

Lastly, it should be noted that outside the data, the GP reverts to
the prior, and consequently the choice of the prior can also play a
significant role in determining the optimal policy inferred by the
SDP.  In the examples shown here we have selected a prior that is both
relatively uninformative (due to the broad priors placed on its parameters
$\ell$ and $\sigma$ and simple (mean zero, radial basis function kernel).
In practice, both the choice of mean and the choice of the covariance
function may be chosen to confer particular biological properties,
as well as more biologically informed priors for $\ell$ and $\sigma$.
In principle, this may allow a manager to improve the performance of the
GPDP approach by adding only enough additional detail as is justified.
For instance, it would be possible to use a linear or a Ricker-shaped
mean in the prior without making the much stronger assumption that the
Ricker is the structurally correct model [@Sugeno2013a]. Future research
should focus on identifying criteria to ensure the prior and the value function
are chosen appropriately for the problem at hand.
